\documentclass[12pt,letterpaper,titlepage]{article}
\usepackage[latin1]{inputenc}
\usepackage{fullpage}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{url}
\usepackage{tikz}
\usepackage{pgf}
\usetikzlibrary{arrows,automata}
\author{Dimitri Demergis \\ Jason Lee \\ Stephen Lombardi \\ Marcus McCurdy}
\title{CS544 Group 5 \\ Easy Realtime Protocol (ERP)}
\begin{document}

\maketitle

\tableofcontents

\pagebreak

\section{Introduction}
This paper defines a media streaming protocol that allows a properly implemented server to stream media to connected clients. Clients are able the join at any time during the streaming process and may be required to authenticate themselves to the server. The protocol is designed to deal with arbitrary streaming data types, including live or prerecorded video and/or audio. The default port for the server is $32456$.


\section{States}
This section describes the states of the protocol, the messages expected in those states, and the
transitions between states.

\subsection{Disconnected}
The protocol always begins in the Disconnected state.  If server discovery is enabled in the client, 
HERP/DERP messages will be exchanged as long as an available server is found on the network.

The client transitions from the Disconnected to Connecting state immediately after sending the 
server a SessionRequest message.  The server transitions to the Connecting state upon receipt 
of the SessionRequest message.

\subsection{Connecting}
In the Connecting state, the client will wait an implementation dependent amount of time for the 
session message. If the client does not receive the message within the timeout it will resend
the session request to the server.  If the server is already processing a session request for a client
it should ignore subsequent requests.

After the server provisions the session, it sends the client a SessionMessage. This message 
contains the session id the client will use for subsequent communication with the server, as
well as the challenge value for the CHAP authentication, which is described in more detail
in section~\ref{sec:security}.

\subsection{Authenticating}
After the client receives the SessionMessage containing the challenge value, it computes
a hash using the SHA-1 algorithm of the challenge value concatenated with a shared secret. The
client then sends the hash in a ChallengeResponse message to the server. The server compares
the clients hash with its own calculated hash. If the hashes match, then the server begins sending
StreamMessages to the client. If the hash comparison fails, then the server sends the client a
RechallengeMessage that contains a new challenge value. If the client fails an implementation
dependent number of challenges the server will send an AuthenticationError message to the client.

While in the authentication state, the client will remain in the Authentication state until it receives
a RechallengeMessage at which time it will calculate a the new hash and send the 
server a new ChallengeResponse. If the client receives a StreamMessage it will transition to 
the Streaming state. If the client receives no messages before the implementation dependent
timeout expires it will resend the last ChallengeResponse assuming that the packet was lost.

\subsection{Streaming}
Once the server verifies that the authentication hash received from the client matches its own hash value, 
it transitions to the Streaming state and begins sending Streaming messages to the client.  The receipt of 
the first streaming message is an indication to the client that the authentication process terminated 
successfully and that it should also pass to the Streaming state.  

While in this state, the server will send Streaming messages to the client at a periodic rate.  The client
will monitor the sequence numbers of the received packets to detect network transmission problems - if 
these problems pass a certain threshhold, the client can send a Throttle message to the Server, 
requesting that the transmission rate be decreased.  The server can choose to accept or ignore the 
request.  During streaming, the client can also choose to send a Pause message, requesting that the
server pause or unpause transmission.


\section{Messages}
\subsection{Client-to-Server Messages}
\subsubsection{SessionRequestMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent by the client to the server.  It is used to notify the server that it 
		should provision a new streaming session. It also includes the version numbers of the 
		protocol the client supports.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | }
		\hline
		1 byte & 1 byte \\
		\hline
		Message ID &  Version Number \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. \newline Always set to 1 for this message. \\
		\hline
		Version Number & byte & Identifies the latest version of the protocol supported by the client.
						\newline 1 = Version 1.0 \\
		\hline
	\end{tabular}
	\end{description}
\subsubsection{ChallengeResponseMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent from the client to the server as part of the authentication process.  
		Upon receiving the ChallengeMessage, the client passes the received challenge value 
		(as well as its own password) to a hash function and sends the result back to the server.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | c | }
		\hline
		1 byte & 1 byte & 1 byte & n bytes \\
		\hline
		Message ID & Session ID & Response length &  Challenge Response \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm}  | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 4 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
		\hline
		Response length & byte & The length of the response in bytes \\
		\hline
		Challenge Response & bytes & Contains the client's challenge response. \\
		\hline
	\end{tabular}
	\end{description}
\subsubsection{DisconnectMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent from the client to the server to indicate that the connection should be closed.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | }
		\hline
		1 byte & 1 byte \\
		\hline
		Message ID & Session ID \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 6 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
		\hline
	\end{tabular}
	\end{description}
	
	\subsubsection{HERPMessage}
	\begin{description}
	\item[Description] \hfill \\
		The Hello Easy Realtime Protocol (HERP) message is broadcast from the client to attempt to auto discover a server on the local network.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | }
		\hline
		1 byte \\
		\hline
		Message ID \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 12 for this message. \\
		\hline
	\end{tabular}
	\end{description}
	
\subsubsection{ThrottleMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent from the client to the server to control the rate of data being sent to the client. It can be used to increase or decrease rate of transmission.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | }
		\hline
		1 byte & 1 byte & 4 bytes \\
		\hline
		Message ID & Session ID & Rate \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 7 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
		\hline
        Rate & float & Indicates the client's desired rate of transmission in bytes per second. \\
        \hline
	\end{tabular}
	\end{description}
\subsubsection{PauseMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent from the client while in the Streaming state to request the server \\
		to Pause or Unpause transmission of Streaming messages.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | }
		\hline
		1 byte & 1 byte & 1 byte \\
		\hline
		Message ID & Session ID & Pause Indication \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 11 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
		\hline
        		Pause Indication & byte & Indicates the desired Pause state.
					\newline 0 = Unpause
					\newline 1 = Pause \\
        		\hline
	\end{tabular}
	\end{description}

\subsection{Server-to-Client Messages}
\subsubsection{SessionMessage}
	\begin{description}
	\item[Description] \hfill \\
		This is the server's reply to the SessionRequestMessage. It tells the client what version of 
		the protocol that will be used during the streaming session, and contains the session id 
		that will be used in further communication with the server.  It also contains the first 
		authentication challenge value sent from the server to the client, as well as the parameters 
		the audio file that the server intends to stream.
		This message transitions the protocol from the Connecting state to the Authenticating state.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | c | c | c |}
		\hline
		1 byte & 1 byte & 1 byte & 4 bytes & 4 bytes & 4 bytes  \\
		\hline
		Message ID & Session ID & Version Number & Chal. Value & Sample Rate & Sample Size  \\
		\hline
	\end{tabular}
	\item \hfill \\
	\begin{tabular}{ | c | c | c |}
		\hline
		4 bytes & 1 bit & 1 bit \\
		\hline
		Num. of Channels   & Signed Flag  & Big-Endian Flag \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. \newline Always set to 2 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this connection. \\
		\hline
		Version Number & byte & Identifies the protocol version to be used in this connection.
						\newline 1 = Version 1.0 \\
		\hline
		Challenge Value & int & Contains the server-generated challenge value.  \\
		\hline
		Sample Rate & float & Contains the audio file's sampling rate (samples per second).  \\
		\hline
		Sample Size & int & Contains the number of bits in each sample.  \\
		\hline
		Channels & int & Contains the number of channels in the audio stream.  \\
		\hline
		Signed Flag & bool & Indicates whether the audio data is signed. 
						\newline 0 = Unsigned
						\newline 1 = Signed \\
		\hline
		Big-Endian Flag & bool & Indicates whether bytes of audio data are in big-endian order. 
						\newline 0 = Little-endian
						\newline 1 = Big-endian \\
		\hline
	\end{tabular}
	\end{description}
	
		\subsubsection{DERPMessage}
	\begin{description}
	\item[Description] \hfill \\
		The Discovered Easy Realtime Protocol message is sent from the server as a response to the clients broadcast HERP message.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | }
		\hline
		1 byte & 4 bytes\\
		\hline
		Message ID & Server Port \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 12 for this message. \\
		\hline
		Port & int & Contains the port number the server is using. \\
		\hline
	\end{tabular}
	\end{description}
	
\subsubsection{RechallengeMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent from the server to the client to indicate an invalid Challenge Response 
		causing an authentication failure.  The message contains a new challenge value generated by the server.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | }
		\hline
		1 byte & 1 byte & 4 bytes \\
		\hline
		Message ID & Session ID &  Challenge Value \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 3 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
		\hline
		Challenge Value & int & Contains the server-generated challenge value. \\
		\hline
	\end{tabular}
	\end{description}

%\subsubsection{ChallengeResultMessage}
%	\begin{description}
%	\item[Description] \hfill \\
%		This message is sent from the server to the client to complete the authentication process.  
%		Upon receiving the ChallengeResponseMessage, the server compares the result received from 
%		the client to its own result, and informs the client of success or failure.
%	\item[Message Format] \hfill \\
%	\begin{tabular}{ | c | c | c | }
%		\hline
%		1 byte & 1 byte & 1 byte \\
%		\hline
%		Message ID & Session ID &  Challenge Result \\
%		\hline
%	\end{tabular}
%	\item[Message Field Definitions] \hfill \\
%	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
%		\hline
%		Field Name & Data Type & Definition \\
%		\hline
%		Message ID & byte & Contains the Message ID number. 
%					\newline Always set to 8 for this message. \\
%		\hline
%		Session ID & byte & Contains the Session ID number pertaining to this message. \\
%		\hline
%		Challenge Result & byte & Contains the result of the challenge process.
%						\newline 0 = Failure 
%						\newline 1 = Success \\
%		\hline
%	\end{tabular}
%	\end{description}

\subsubsection{AuthenticationErrorMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message is sent from the server to the client to notify the client of an error in 
		completing the authentication process. This message contains an error code in order to 
		properly inform the client of the nature of the error. This message transitions the protocol
		state from Authenticating to Disconnected.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | }
		\hline
		1 byte & 1 byte & 4 bytes \\
		\hline
		Message ID & Session ID & Error Code \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 9 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
		\hline
		Error Code & int & Contains the code for the error that occurred during the authentication process. \\
		\hline
	\end{tabular}
	\end{description}	
\subsubsection{StreamMessage}
	\begin{description}
	\item[Description] \hfill \\
		These messages are sent while the protocol is in the streaming state. They contain the 
		actual data that is being streamed. The message consists of a message ID identifying it as a data message, a session ID identifying the session, a four-byte data length field, and a variable $n$-byte data chunk.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | c | c | }
		\hline
		1 byte & 1 byte & 1 byte & 4 bytes & $n$ bytes \\
		\hline
		Message ID & Session ID & Sequence Number & Data Length & Encoded Data \\
		\hline
	\end{tabular}
	\item \hfill \\
	\begin{tabular}{ | c | c | }
		\hline
		4 bytes & $m$ bytes \\
		\hline
		CRC length & CRC \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
					\newline Always set to 5 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
        \hline
        Sequence Number & byte & Sequence number for this message. \\
		\hline
		Data Length & int & Contains the value of $n$. \\
		\hline
        Encoded Data & variable & Encoded stream data. \\
        \hline
        CRC Length & int & The length of the CRC. \\
        \hline
        CRC & byte & Cyclic redundancy check for the data. \\
        \hline
	\end{tabular}
	\end{description}
	
	\subsubsection{StreamErrorMessage}
	\begin{description}
	\item[Description] \hfill \\
		This message may be sent while the protocol is in the streaming state. It signifies an unrecoverable
		error during the streaming process. It causes the protocol to transition from the streaming to
		disconnected state.
	\item[Message Format] \hfill \\
	\begin{tabular}{ | c | c | c | }
		\hline
		1 byte & 1 byte & 4 bytes \\
		\hline
		Message ID & Session ID & Error Code \\
		\hline
	\end{tabular}
	\item[Message Field Definitions] \hfill \\
	\begin{tabular}{ | p{3cm} | p{1.5cm} | p{8cm} | }
		\hline
		Field Name & Data Type & Definition \\
		\hline
		Message ID & byte & Contains the Message ID number. 
		\newline Always set to 10 for this message. \\
		\hline
		Session ID & byte & Contains the Session ID number pertaining to this message. \\
	    \hline
		Error Code & int & Contains the code for the error that occurred during streaming. \\
		\hline
	\end{tabular}
	\end{description}

\section{DFA}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,bend angle=75,inner sep=1pt, every state/.style={draw=blue!50,very thick,fill=blue!20, font=\footnotesize}]

\foreach \name/\text/\angle in {1/disconnected/180, 2/connecting/90,  3/authenticating/0, 4/streaming/270}
\node[state,xshift=10cm,yshift=2cm](\name) at (\angle:5cm) {\text};

\path 
	(1) edge				node {session req} 		(2)
	(2) edge				node {session msg} 		(3)
	      edge [loop right]		node {timeout} 			(2)
	      edge [loop above]	node{session req}		(2)	
	(3)	
	      edge [loop right]		node {rechallenge}          	(4)
	      edge 				node {auth err}          		(1)
	      edge				node {stream}			(4)
%	      edge [bend left]		node {chlg result}		(4)
	(4) edge				node {disconnect}		(1)
	      edge [loop below]	node {throttle}			(4)
	      edge [loop above] node{stream}			(4)
	      edge [loop right] node{pause}			(4)
	      edge [bend left]		node {stream err}		(1)	      	
	
	;

\end{tikzpicture}

%session request id = 1
% ChallengeResponseMessage = 4
% Disconnect = 6
% Throttle = 7
%SessionMessage = 2
%Pause = 3
% ChallengeResult = 8
%auth err = 9
%stream = 5
%rechallenge


\section{Extensibility}
Protocol version information is included in connection messages so that the protocol version used by both sides can be established during the configuration phase.  This allows future improvements to the protocol to be implemented without rendering existing client applications unusable.

\section{Security}
\label{sec:security}
This protocol uses a CHAP-like handshaking protocol for authentication of clients.  After the connection has been established, the server sends a challenge message to the client.  Both the client and server pass a password through a pre-determined hash function, and the server checks the results and informs the client of success or failure. For this implementation, the shared secret
is hardcoded to "CS544GROUP5". This would correspond to a users password in a full blown
application.

\section{Reliability}
Due to the real-time nature of the protocol, it does not attempt to retransmit any lost packets. It does, however, provide sequence number on all data messages in order to ignore any packets which have reached the client after a more recently sent packet. In addition to this, each data message contains a CRC for basic error checking. When an error is detected, the packet is simply dropped.

\section{Quality of Service}
Streaming media transmission requires careful monitoring of traffic to maintain acceptable levels of loss and to avoid network congestion. These goals are achieved through the ThrottleMessage. The client sends a ThrottleMessage to inform the server that it would like to change the server's rate of transmission to that client. If the network is under a heavy load, the client will observe that packets have been lost in transit by examining sequence numbers. When this occurs, the client will request that the server decrease its current transmission rate to avoid packet loss for other services across the network. This is necessary because UDP will not perform this operation on its own like TCP would. If the client wishes to increase the rate of transmission it may do this as well, but the server is free to ignore any request that is too high.

\section{Auto Discovery}
\label{sec:autodiscovery}
The client will auto discover local servers by sending out a HERP message to the broadcast address. It will then wait 1 second and if it does not hear a reply it will send out another broadcast. It will try up to 5
times before giving up. When the server receives the HERP message it creates a DERP message that it send back to the address and port that the HERP message was sent from. The client then utilizes the information in the
HERP message to begin the protocol. The auto discovery implementation listens on port 65530,
which is different than the standard server port of 32456. This was purely an implementation 
choice to keep the discovery isolated from the main code in case it caused any issues. 

This approach should work as long as the client and server are on the same subnet. It is possible that it will work if they are not, but that would be dependent on whether or not the router on the network will route a
a UDP packet sent to the broadcast address 255.255.255.255.

\section{Changes from Original Proposal}
\subsection{Restructuring of DFA}
During implementation, we discovered that two of the protocol's states were unnecessary and have removed them:
\begin{itemize}
\item {\bf Connected State - }Upon sending the Session message to the client, the server would transition to Connected, only to transition to Authenticating moments later after sending the first Challenge message.
\item {\bf Authenticated State - }Upon sending the Challenge Result message to the client, the server would transition to Authenticated, only to transition to Streaming moments later after sending the first Streaming message.
\end{itemize}

\subsection{Authentication Changes}
In the original design, upon receiving the SessionRequest message from the client, the server would respond with a SessionMessage, immediately followed by the first Challenge message.  This created a race condition whereby the client might receive the Challenge before receiving the the SessionMessage.  As a result, the SessionMessage has been modified to include the challenge value for the first challenge.  The Challenge message has been renamed Rechallenge, as it is now only used when the first challenge fails and a new challenge value must be sent to the client. 

Additionally, the Challenge Result message has been removed, since the above changes have made it redundant.  In case of challenge failure, the server sends Rechallenge with a new challenge value.  In case of challenge success, the server sends the first Streaming message to signal the transition to the Streaming state.

\subsection{Server Discovery}
Two new messages (HERP and DERP) have been added to the protocol to allow the client to search for available servers without providing a specific IP address.  See section ~\ref{sec:autodiscovery} above for more information.

\subsection{Audio Format Parameters}
The SessionMessage has been updated to contain separate fields for all parameters that the client needs to properly process the audio file being streamed by the server, rather than sending all this information in a single variable-length ASCII-encoded string field.

\subsection{Pause Feature}
Pause is a new feature that has been added since the original design.  The user can enter 'p' to send a Pause message to the server indicating that it should temporarily pause streaming.  Pressing 'p' again sends another message to the server indicating Unpause.

\section{Performance}
The implementation of the protocol is only streaming uncompressed audio in the WAV file format.
As such, the playback of the audio can become choppy if the file being streamed has a high
sample rate and multiple channels. It would be desirable to stream a compressed file format, but
due to time constraints we did not write any code to perform audio compression.

\section{Sample Run}
A sample run of the client can be found here \url{http://www.youtube.com/watch?v=nw3SDh7rij8}

\section{Robustness}
The implementation is fairly robust against fuzzing. The client and server check the message
type of every message and then their current state. Both will drop messages of unexpected type based on their current state. In addition, every message is tested for proper session id and is
dropped if there is a mismatch.

Something that the protocol could be vulnerable to is an adversary sending datagrams to
the client's socket that contain almost valid protocol messages. It is possible that the client
could receive a packet where the first byte in the data of that packet corresponds to one of the
message ids of a valid message. However, the client will attempt to read other data from the packet
and most likely fail, but it will then just drop that packet and should continue normal operation.

\section{Group Contribution}
Everyone in the group contributed to this writeup. Each member contributed one or more states and messages as well as writing sections and subsections of this document. Everyone provided valuable feedback and ideas during this initial design phase of the protocol.


Everyone in the group committed code to the assignment as well. It as a group effort. All members
contributed to both the client, server, and shared code used on the project.


\end{document}
